### 2015-05-12 background.tex
fix

Nowadays, multicore machines have become mainstream, allowing operating systems to exploit concurrency for \emph{responsiveness} (latency) or \emph{performance} (throughput). In drivers, responsiveness is important, because even if there is only a single core, entry points of distinct processes should run concurrently so that no single process hogs the core. Likewise, performance is also important: e.g. network drivers should be able to process incoming packages in a timely manner.

@@

The most common technique for eliminating data races is using a \emph{locking} mechanism. Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e. source code that cannot be executed by more than one thread simultaneously. Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in Figure~\ref{fig:data_race_example}. The example is tricky, because the return statement can potentially race on the \texttt{f\_pos} field. To fix this issue, we store the result in a temporary variable, inside the critical section.

Lock-based synchronization is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}. Coarse-grained locking can hurt performance, because it limits the opportunity for concurrency. On the other hand, fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read and write operations, in this work we focus on locks. However, we still treat lock-free operations (e.g. atomics) soundly (see Definition~\ref{definition:datarace}), by not regarding them as providing any protection between threads. This is sound but can potentially introduce false positives.

>>>>

Nowadays, multicore machines have become mainstream, allowing operating systems to exploit concurrency for \emph{responsiveness} (latency) or \emph{performance} (throughput). In drivers, responsiveness is important (even if there is only a single core), because entry points of distinct processes should run concurrently so that no single process hogs the core. Likewise, performance is also important: e.g.\ network drivers should be able to process incoming packages in a timely manner.

@@

The most common technique for eliminating data races is using a \emph{locking} mechanism. Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e.\ source code that cannot be executed by more than one thread simultaneously. Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in Figure~\ref{fig:data_race_example}. The example is tricky, because the return statement can potentially race on the \texttt{f\_pos} field. To fix this issue, we store the result in a temporary variable, inside the critical section.

Lock-based synchronization is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}. Coarse-grained locking can hurt performance, because it limits the opportunity for concurrency. On the other hand, fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read and write operations, in this work we focus on locks. However, we still treat lock-free operations (e.g.\ atomics) soundly (see Definition~\ref{definition:datarace}), by not regarding them as providing any protection between threads. This is sound but can potentially introduce false positives.

### 2015-05-12 intro.tex
fix

Techniques such as~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been successfully used to analyze device drivers, but mostly focus on generic sequential program properties and finding protocol bugs. Well-regarded Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races. Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e. can miss real bugs) or \emph{imprecise} (i.e. can report false bugs), and typically sacrifice precision for scalability. Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.

@@

\whoop currently supports only Linux drivers, but the approach is generic enough and thus we believe that it can be used to analyze drivers from other operating systems (e.g. Windows) as well as other concurrent programs with similar structure (e.g. filesystems).

>>>>

Techniques such as~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been successfully used to analyze device drivers, but mostly focus on generic sequential program properties and finding protocol bugs. Well-regarded Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races. Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability. Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.

@@

\whoop currently supports only Linux drivers, but the approach is generic enough and thus we believe that it can be used to analyze drivers from other operating systems (e.g.\ Windows) as well as other concurrent programs with similar structure (e.g.\ filesystems).

### 2015-05-12 whoop.tex
fix

Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency offered by the \corral tool, based on context bounding and stratified inlining; we discuss the latter approach in Section~\ref{whoop:bugfinding}.

@@

Figure~\ref{fig:whoop} depicts the \whoop toolchain. Initially, \whoop accepts a Linux driver written in C, together with a Linux environmental model\footnote{Stub header files that model the Linux kernel APIs.}, which is required to ``close'' the driver and allow it to be subsequently over-approximated and analyzed. Both the driver and the environmental model are passed to the translation phase of \whoop (see Section~\ref{whoop:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}. Next, \whoop performs \emph{static pairwise lockset analysis} to detect all potential data races in the abstract program. The proof sketch behind our static lockset analysis is given in Section~\ref{whoop:proof}. The instrumentation and analysis is described in more detail in Section~\ref{whoop:method}. The summarization that we perform to achieve scalability is presented in Section~\ref{whoop:summarization}. After the static lockset analysis phase, \whoop exploits the race-related information to accelerate bug-finding with \corral (see Section~\ref{whoop:bugfinding})

@@

First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given driver, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points. This function acts as the main entry point to the driver after the latter has been registered with the kernel. This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{whoop:method} and~\ref{whoop:bugfinding}). The output is a file that contains information regarding the entry points and a file with the rewritten program.

@@

Statically analyzing system-level software, such as drivers, requires to ``close'' the environment by abstracting away the low level implementation details. In this work, we developed a simple environmental model for the Linux kernel that consists of (nondeterministic) stub functions. A limitation of our model is that it can, and will, ultimately result in decreased precision and, thus, false positives. However, because we currently only focus on finding data races, we can get away with over-approximating a lot of the underlying Linux kernel functionality, without losing too much precision. Enriching our model with domain-specific knowledge to make it more precise is an ongoing manual effort, but requires Linux expertise. We argue that further increasing the precision is orthogonal to the contributions of this paper. Moreover, even if the symbolic lockset analysis results in false positives, \whoop can still use any race-freedom guarantees to significantly speedup a more precise bug-finder, as seen in Sections~\ref{whoop:bugfinding} and~\ref{evaluation}.

>>>>

Our experimental evaluation (Section~\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency offered by the \corral tool, based on context bounding and stratified inlining; we discuss the latter approach in Section~\ref{corral}.

@@

Figure~\ref{fig:whoop} depicts the \whoop toolchain. Initially, \whoop accepts a Linux driver written in C, together with a Linux environmental model\footnote{Stub header files that model the Linux kernel APIs.}, which is required to ``close'' the driver and allow it to be subsequently over-approximated and analyzed. Both the driver and the environmental model are passed to the translation phase of \whoop (see Section~\ref{whoop:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}. Next, \whoop performs \emph{static pairwise lockset analysis} to detect all potential data races in the abstract program. The proof sketch behind our static lockset analysis is given in Section~\ref{whoop:proof}. The instrumentation and analysis is described in more detail in Section~\ref{whoop:method}. The summarization that we perform to achieve scalability is presented in Section~\ref{whoop:summarization}. After the static lockset analysis phase, \whoop exploits the race-related information to accelerate bug-finding with \corral (see Section~\ref{corral})

@@

First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given driver, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points. This function acts as the main entry point to the driver after the latter has been registered with the kernel. This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{whoop:method} and~\ref{corral}). The output is a file that contains information regarding the entry points and a file with the rewritten program.

@@

Statically analyzing system-level software, such as drivers, requires to ``close'' the environment by abstracting away the low level implementation details. In this work, we developed a simple environmental model for the Linux kernel that consists of (nondeterministic) stub functions. A limitation of our model is that it can, and will, ultimately result in decreased precision and, thus, false positives. However, because we currently only focus on finding data races, we can get away with over-approximating a lot of the underlying Linux kernel functionality, without losing too much precision. Enriching our model with domain-specific knowledge to make it more precise is an ongoing manual effort, but requires Linux expertise. We argue that further increasing the precision is orthogonal to the contributions of this paper. Moreover, even if the symbolic lockset analysis results in false positives, \whoop can still use any race-freedom guarantees to significantly speedup a more precise bug-finder, as seen in Sections~\ref{corral} and~\ref{evaluation}.
